---
title: "part2_H3K27ac"
output: html_document
date: "2026-06-23"
---

# Set up

```{r}
suppressPackageStartupMessages({
  library(edgeR)
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT)
  library(SummarizedExperiment)
  library(EnrichedHeatmap)
  library(ComplexHeatmap)
  library(AnnotationHub)
  library(MotifDb)
  library(memes)
  library(universalmotif)
  library(ensembldb)
  library(motifmatchr)
  library(sechm)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(limma)
  library(betterChromVAR)
})
```

```{r}
count_matrix <- readRDS("peaks/h3k27ac_count_matrix.rds")
consensus <- readRDS("peaks/H3K27ac_consensus.rds")
tracks <- list.files(path = "peaks", pattern="bw$", full.names = TRUE)
names(tracks) <- gsub("\\.bw", "", basename(tracks))
```

```{r}
h3k27ac_signal <- signal2Matrix(tracks, consensus, extend=2000)
```

# Plotting all replicates

```{r, fig.height=5, fig.width=15}
plotEnrichedHeatmaps(h3k27ac_signal)
```

The heatmap shows H3K27ac signal for all samples and replicates separately. Heatmap intensity represents RPKM-normalized H3K27ac coverage around consensus regions. Replicates seem to agree with each other except for those originating from coculture in hypoxic conditions.

# Averaging per sample

```{r}
average_matrices <- function(mats, new_name) {
  avg <- Reduce("+", lapply(mats, as.matrix)) / length(mats)
  
  attributes(avg) <- attributes(mats[[1]])
  attr(avg, "signal_name") <- new_name
  class(avg) <- class(mats[[1]])
  
  avg
}
```

```{r}
mat_list <- as.list(assay(h3k27ac_signal, "input"))

Ctrl_N_mat <- average_matrices(
  mat_list[c("H3K27ac_Ctrl_N_1", "H3K27ac_Ctrl_N_2")],
  "Ctrl_N"
)

Ctrl_H_mat <- average_matrices(
  mat_list[c("H3K27ac_Ctrl_H_1", "H3K27ac_Ctrl_H_2")],
  "Ctrl_H"
)

CC_N_mat <- average_matrices(
  mat_list[c("H3K27ac_CC_N_1", "H3K27ac_CC_N_2")],
  "CC_N"
)

CC_H_mat <- average_matrices(
  mat_list[c("H3K27ac_CC_H_1", "H3K27ac_CC_H_2")],
  "CC_H"
)
```

```{r}
avg_mats <- list(CC_H = CC_H_mat, CC_N = CC_N_mat, Ctrl_H = Ctrl_H_mat, Ctrl_N = Ctrl_N_mat)
plotEnrichedHeatmaps(avg_mats)
```

Loss of acetylation in hypoxia / co-culture / hypoxia + co-culture

# Differential analysis

```{r}
# Define groups based on the column names of the count matrix 
if (!identical(colnames(count_matrix), c("Ctlr_N_1", "Ctlr_N_2", "Ctlr_H_1", "Ctlr_H_2", "CC_N_1", "CC_N_2", "CC_H_1", "CC_H_2"))) {
  stop("Column names do not match.")
}
groups <- factor(c("Ctrl_N", "Ctrl_N", "Ctrl_H", "Ctrl_H", "CC_N", "CC_N", "CC_H", "CC_H"))

# Initialize edgeR object, filter background noise, and normalize
y <- DGEList(counts = count_matrix, group = groups)
keep <- filterByExpr(y)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

# Set up the statistical design
design <- model.matrix(~0 + groups)
colnames(design) <- levels(groups)

# Estimate variance (dispersion) and fit the model
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design) # Quasi-Likelihood Negative Binomial

# Define exact experimental questions
my_contrasts <- makeContrasts(   
  Ctrl_H_vs_Ctrl_N = Ctrl_H - Ctrl_N, # Baseline hypoxia response 
  CC_N_vs_Ctrl_N  = CC_N - Ctrl_N,   # Baseline co-culture response
  CC_H_vs_CC_N   = CC_H - CC_N,     # Hypoxia effect during co-culture
  levels = design
)

# Run the tests for all contrasts
test1 <- glmQLFTest(fit, contrast=my_contrasts[,"Ctrl_H_vs_Ctrl_N"])
test2 <- glmQLFTest(fit, contrast=my_contrasts[,"CC_N_vs_Ctrl_N"])
test3 <- glmQLFTest(fit, contrast=my_contrasts[,"CC_H_vs_CC_N"])

# Extract the significant tables for each
diff_Ctrl_H_vs_Ctrl_N <- topTags(test1, n=Inf, p.value=1, sort.by = "none")$table
diff_CC_N_vs_Ctrl_N <- topTags(test2, n=Inf, p.value=1, sort.by = "none")$table
diff_CC_H_vs_CC_N <- topTags(test3, n=Inf, p.value=1, sort.by = "none")$table

head(diff_Ctrl_H_vs_Ctrl_N)

# logFC > 0 = more acetylated / stronger active regulatory mark
```

## Plot volcano plot of DA peaks
```{r}
plot_volcano <- function(df, plot_title) {
  # Categorize each peak
  df$Significance <- "Not Significant"
  df$Significance[df$FDR < 0.05 & df$logFC > 0.6] <- "Up"
  df$Significance[df$FDR < 0.05 & df$logFC < -0.6] <- "Down"
  
  # Plotting volcano
  p <- ggplot(df, aes(x = logFC, y = -log10(FDR), color = Significance)) +
    geom_point(alpha = 0.6, size = 1.2) +
    scale_color_manual(values = c("Down" = "blue", "Not Significant" = "gray", "Up" = "red")) +
    theme_minimal() +
    labs(title = plot_title, x = "Log2 Fold Change", y = "-Log10 FDR") +
    geom_vline(xintercept = c(-0.6, 0.6), linetype = "dashed") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed")
  
  print(p)
  return(df)
}

diff_Ctrl_H_vs_Ctrl_N <- plot_volcano(diff_Ctrl_H_vs_Ctrl_N, "Hypoxia vs Normoxia in Monoculture")
diff_CC_H_vs_CC_N <- plot_volcano(diff_CC_H_vs_CC_N, "Hypoxia vs Normoxia in Co-Culture")
diff_CC_N_vs_Ctrl_N <- plot_volcano(diff_CC_N_vs_Ctrl_N, "Co vs Monocolture in Normoxia")
```

```{r}
# importing differential accessibility results from ATAC-seq analysis
accessibility_diff_Ctrl_H_vs_Ctrl_N <- readRDS("diff_Ctrl_H_vs_Ctrl_N.rds")
accessibility_diff_CC_N_vs_Ctrl_N   <- readRDS("diff_CC_N_vs_Ctrl_N.rds")
accessibility_diff_CC_H_vs_CC_N     <- readRDS("diff_CC_H_vs_CC_N.rds")
accessibility_consensus_kept        <- readRDS("accessibility_consensus_kept.rds")
```

```{r}
count_sig <- function(df, name) {
  up   <- sum(df$Significance == "Up")
  down <- sum(df$Significance == "Down")
  total <- sum(up, down)
  cat(name, "— Up:", up, "| Down:", down, "| Total: ", total, "\n")
}

count_sig(diff_Ctrl_H_vs_Ctrl_N, "Ctrl_H vs Ctrl_N")
count_sig(diff_CC_N_vs_Ctrl_N,   "CC_N vs Ctrl_N")
count_sig(diff_CC_H_vs_CC_N,     "CC_H vs CC_N")
```

# Comparing changes in acetylation to changes in accessibility

```{r}
# filtering consensus sequences to keep only those that are relevant for differential acetylation analysis
if (!identical(rownames(count_matrix), names(consensus))) {
  stop("Row order does not match between count_matrix and consensus.")
}
consensus_kept <- consensus[keep]

# reordering accessibility_consensus_kept
accessibility_consensus_kept <- accessibility_consensus_kept[
  rownames(accessibility_diff_CC_H_vs_CC_N)
]

# This function takes the differential analysis output table and attaches its results to the respective genomic coordinates
make_diff_granges <- function(edgeR_table, consensus_kept) {
  gr <- consensus_kept
  
  mcols(gr)$logFC  <- edgeR_table$logFC
  mcols(gr)$logCPM <- edgeR_table$logCPM
  mcols(gr)$PValue <- edgeR_table$PValue
  mcols(gr)$FDR    <- edgeR_table$FDR
  
  gr
}

# converting the differential analysis output to GRanges
acetylation_Ctrl_H_vs_Ctrl_N <- make_diff_granges(
  diff_Ctrl_H_vs_Ctrl_N,
  consensus_kept
)

acetylation_CC_N_vs_Ctrl_N <- make_diff_granges(
  diff_CC_N_vs_Ctrl_N,
  consensus_kept
)

acetylation_CC_H_vs_CC_N <- make_diff_granges(
  diff_CC_H_vs_CC_N,
  consensus_kept
)

accessibility_CC_H_vs_CC_N <- make_diff_granges(
  accessibility_diff_CC_H_vs_CC_N,
  accessibility_consensus_kept
)
```

```{r}
acetylation_Ctrl_H_vs_Ctrl_N <- acetylation_Ctrl_H_vs_Ctrl_N[acetylation_Ctrl_H_vs_Ctrl_N$FDR < 0.05 & abs(acetylation_Ctrl_H_vs_Ctrl_N$logFC) > 0.6]
acetylation_CC_N_vs_Ctrl_N <- acetylation_CC_N_vs_Ctrl_N[acetylation_CC_N_vs_Ctrl_N$FDR < 0.05 & abs(acetylation_CC_N_vs_Ctrl_N$logFC) > 0.6]
acetylation_CC_H_vs_CC_N <- acetylation_CC_H_vs_CC_N[acetylation_CC_H_vs_CC_N$FDR < 0.05 & abs(acetylation_CC_H_vs_CC_N$logFC) > 0.6]

accessibility_CC_H_vs_CC_N <- accessibility_CC_H_vs_CC_N[accessibility_CC_H_vs_CC_N$FDR < 0.05 & abs(accessibility_CC_H_vs_CC_N$logFC) > 0.6]
```

```{r, fig.height=6, fig.width=7}
# adding group labels:
acetylation_CC_H_vs_CC_N$group = ifelse(acetylation_CC_H_vs_CC_N$logFC > 0, "more acetylated in CCH", "more acetylated in CCN")
accessibility_CC_H_vs_CC_N$group = ifelse(accessibility_CC_H_vs_CC_N$logFC > 0, "more accessible in CCH", "more accessible in CCN")

regions <- list(Acetylation_with_hypoxia_in_CC = acetylation_CC_H_vs_CC_N[acetylation_CC_H_vs_CC_N$group == "more acetylated in CCH"], 
                Deacetylation_with_hypoxia_in_CC = acetylation_CC_H_vs_CC_N[acetylation_CC_H_vs_CC_N$group == "more acetylated in CCN"], 
                Accessibility_with_hypoxia_in_CC = accessibility_CC_H_vs_CC_N[accessibility_CC_H_vs_CC_N$group == "more accessible in CCH"],
                Inaccessibility_with_hypoxia_in_CC = accessibility_CC_H_vs_CC_N[accessibility_CC_H_vs_CC_N$group == "more accessible in CCN"])

# visualizing overlaps between accessible/acetylated regions using a heatmap of overlaps:
regionOverlaps(regions)
```

The relationship between changes in H3K27ac and changes in accessibility is not clear-cut: we can see that loss of accessibility is linked to deacetylation at H3K27, whereas increased accessibility is linked to both acetylation and deacetylation (acetylation + increased accessibility is a more common combination though).

# H3K27ac signal around enhancer regions

Active enhancer peaks are OPEN and HAVE h3k27ac histone modifications, poised enhancers are OPEN but do NOT HAVE h3k27ac modifications, and inactive enhancers are CLOSED and also do NOT HAVE h3k27ac modifications.

```{r}
# importing enh_up and enh_down from part2
enh_down <- readRDS("enh_down.rds")
enh_up   <- readRDS("enh_up.rds")
```

```{r}
extract_signal <- function(regions_of_interest, tracks, extend=2000, w = NULL) {
  if (!(class(regions_of_interest) == "GRanges")) {
    regions_of_interest <- makeGRangesFromDataFrame(regions_of_interest, keep.extra.columns=TRUE)
  }
  
  # we extract signals around regions of interest
  regions_of_interest_signal <- signal2Matrix(tracks, regions_of_interest, extend=extend, w=w)
  
  # averaging between replicates
  mat_list <- as.list(assay(regions_of_interest_signal, "input"))
  
  CC_N_mat <- average_matrices(mat_list[c("H3K27ac_CC_N_1", "H3K27ac_CC_N_2")], "CC_N")
  
  CC_H_mat <- average_matrices(mat_list[c("H3K27ac_CC_H_1", "H3K27ac_CC_H_2")], "CC_H")
  
  avg_mats <- list(CC_H = CC_H_mat, CC_N = CC_N_mat)
  
  return(avg_mats)
  }
```

```{r}
plotEnrichedHeatmaps(extract_signal(enh_up, tracks), trim=0.95, multiScale = FALSE, axis_name = "0")
plotEnrichedHeatmaps(extract_signal(enh_down, tracks), trim=0.95, multiScale = FALSE, axis_name = "0")
```

1st plot: accessibility of enhancer candidates increases in CC_H compared to CC_N, but H3K27ac signal tends to decrease. We would expect at least a similar if not higher acetylation level (heatmap indicates that accessibility is often accompanied by acetylation, why not here?) 

2nd plot: accessibility of enhancer candidates decreases in CC_H compared to CC_N, and so does H3K27ac signal. Explanation -> enhancers becoming inactive?

```{r}
plotEnrichedHeatmaps(extract_signal(enh_up, tracks), cluster_rows = TRUE, show_row_dend=TRUE)
```

I thought we might see subpopulations where acetylation increases with accessibility but that doesn't seem to be the case.

# H3K27ac signal around genic regions

```{r}
# importing genic_up and genic_down from part2
genic_up   <- readRDS("genic_up.rds")
genic_down <- readRDS("genic_down.rds")
```

```{r}
plotEnrichedHeatmaps(extract_signal(genic_up, tracks), trim=0.95, multiScale = FALSE, axis_name = "0")
```

```{r}
plotEnrichedHeatmaps(extract_signal(genic_down, tracks), trim=0.95, multiScale = FALSE, axis_name = "0")
```

# H3K27ac signal around TF motif occurences

Here we investigate the TFs SPI1, IRF8 (enhancer DOWN) and NF-κB (enhancer UP)

```{r}
get_motifs <- function(motif_name, regions_of_interest, reference = "Mus", genome = BSgenome.Mmusculus.UCSC.mm10) {
  if (!(class(regions_of_interest) == "GRanges")) {
    regions_of_interest <- makeGRangesFromDataFrame(regions_of_interest, keep.extra.columns=TRUE)
  }
  
  # obtaining the motif sequence  
  motif_seq <- MotifDb::query(MotifDb, c(motif_name, reference))[[1]]
  motif_seq_converted <- convert_motifs(motif_seq, class="TFBSTools-PFMatrix")
  
  # find motif matches across background
  motif <- motifmatchr::matchMotifs(pwms = motif_seq_converted, subject = regions_of_interest, genome = genome, out="positions", p.cutoff=1e-5)[[1]]
  return(motif)
}
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_SPI1 <- extract_signal(get_motifs("SPI1", enh_down), tracks, w=5) 
avg_mats_IRF8 <- extract_signal(get_motifs("IRF8", enh_down), tracks, w=5) 
avg_mats_NFKB <- extract_signal(get_motifs("NFKB1", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_SPI1, trim=0.95, multiScale = FALSE, axis_name = "0")
plotEnrichedHeatmaps(avg_mats_IRF8, trim=0.95, multiScale = FALSE, axis_name = "0")
plotEnrichedHeatmaps(avg_mats_NFKB, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

Plotting aggregated signal

```{r}
get_profile <- function(mat, motif_name, extend = 2000, w = 5) {
  profile <- colMeans(as.matrix(mat), na.rm = TRUE)
  data.frame(
    position_bp = seq(-extend, extend - w, by = w),
    signal = profile,
    motif = motif_name
  )
}

profile_df <- rbind(
  get_profile(avg_mats_SPI1$CC_H, "SPI1_CC_H"),
  get_profile(avg_mats_IRF8$CC_H, "IRF8_CC_H"),
  get_profile(avg_mats_NFKB$CC_H, "NFKB_CC_H"),
  get_profile(avg_mats_SPI1$CC_N, "SPI1_CC_N"),
  get_profile(avg_mats_IRF8$CC_N, "IRF8_CC_N"),
  get_profile(avg_mats_NFKB$CC_N, "NFKB_CC_N")
)

ggplot(profile_df, aes(x = position_bp, y = signal, color = motif)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "H3K27ac signal around TF motif occurrences",
    x = "Distance from motif center (bp)",
    y = "Mean H3K27ac signal"
  )
```

Here we investigate the TFs NFAC1.1 (enhancer DOWN), FOSL1, FOSL2, JUND, JUN, JUNB, BATF, FOSB (enhancer UP), and KLF4, SP1 (genic UP).
IRF8 disappeared from the motifs.

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_KLF4 <- extract_signal(get_motifs("KLF4", genic_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_KLF4, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_SP1 <- extract_signal(get_motifs("SP1", genic_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_SP1, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_FOSL1 <- extract_signal(get_motifs("FOSL1", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_FOSL1, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_FOSL2 <- extract_signal(get_motifs("FOSL2", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_FOSL2, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_JUND <- extract_signal(get_motifs("JUND", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_JUND, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_JUN <- extract_signal(get_motifs("JUN", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_JUN, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_JUNB <- extract_signal(get_motifs("JUNB", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_JUNB, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_BATF <- extract_signal(get_motifs("BATF", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_BATF, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_FOSB <- extract_signal(get_motifs("FOSB", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_FOSB, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_NFAC1 <- extract_signal(get_motifs("NFAC1", enh_down), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_NFAC1, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

```{r}
# we extract signals around the motif occurences, zoom in to 2000bp around the motif centers, in windows of 5bp
avg_mats_xxx <- extract_signal(get_motifs("xxx", enh_up), tracks, w=5) 

# plot the signals:
plotEnrichedHeatmaps(avg_mats_xxx, trim=0.95, multiScale = FALSE, axis_name = "0")
# multiScale = TRUE because multiScale = FALSE would simply show a global decrease in acetylation signal
```

# Enrichment analysis

Enrichment for H3K27ac gain

```{r}
# H3K27ac gain in hypoxic co-culture
acetylation_gain_CC_H_vs_CC_N <- acetylation_CC_H_vs_CC_N[acetylation_CC_H_vs_CC_N$logFC > 0.6]

# running GREAT-style enrichment analysis on regions that are differentially acetylated in CC_H vs. CC_N
# we use the mouse mm10 genome annotation to link regions to nearby genes/TSSs
# gene_sets="GO:BP" specifies that we test enrichment for Gene Ontology Biological Process terms
enrichment_acetylation_gain_CC_H_vs_CC_N <- great(acetylation_gain_CC_H_vs_CC_N, gene_sets="GO:BP", tss_source="mm10", background=consensus_kept, cores=2)
enrichment_table_acetylation_gain_CC_H_vs_CC_N <- getEnrichmentTables(enrichment_acetylation_gain_CC_H_vs_CC_N)
enrichment_table_acetylation_gain_CC_H_vs_CC_N <- enrichment_table_acetylation_gain_CC_H_vs_CC_N[order(enrichment_table_acetylation_gain_CC_H_vs_CC_N$p_adjust), ]

head(enrichment_table_acetylation_gain_CC_H_vs_CC_N, 15)
ggplot(head(enrichment_table_acetylation_gain_CC_H_vs_CC_N, 15), aes(fold_enrichment, reorder(description, p_adjust),
                        size=observed_region_hits, color=-log10(p_adjust))) + geom_point() + scale_color_viridis_c()
```

Plotting H3K27ac signal in regions that gain H3K27ac in CC_H compared to CC_N 

```{r}
plotEnrichedHeatmaps(extract_signal(acetylation_gain_CC_H_vs_CC_N, tracks), trim=0.95, multiScale = FALSE, axis_name = "0")
```

The increase in H3K27ac signal, if present, is marginal. Therefore, it is not necessarily visible in averaged H3K27ac signal plots. The heatmaps may be dominated by the larger number or stronger signal change of regions that lose H3K27ac, causing the overall visual pattern to show reduced acetylation in CC_H compared with CC_N, even though a subset of regions shows coordinated gains in H3K27ac.

Enrichment for H3K27ac loss

```{r}
# H3K27ac loss in hypoxic co-culture
acetylation_loss_CC_H_vs_CC_N <- acetylation_CC_H_vs_CC_N[acetylation_CC_H_vs_CC_N$logFC < -0.6]

# running GREAT-style enrichment analysis on regions that are differentially acetylated in CC_H vs. CC_N
# we use the mouse mm10 genome annotation to link regions to nearby genes/TSSs
# gene_sets="GO:BP" specifies that we test enrichment for Gene Ontology Biological Process terms
enrichment_acetylation_loss_CC_H_vs_CC_N <- great(acetylation_loss_CC_H_vs_CC_N, gene_sets="GO:BP", tss_source="mm10", background=consensus_kept, cores=2)
enrichment_table_acetylation_loss_CC_H_vs_CC_N <- getEnrichmentTables(enrichment_acetylation_loss_CC_H_vs_CC_N)
enrichment_table_acetylation_loss_CC_H_vs_CC_N <- enrichment_table_acetylation_loss_CC_H_vs_CC_N[order(enrichment_table_acetylation_loss_CC_H_vs_CC_N$p_adjust), ]

head(enrichment_table_acetylation_loss_CC_H_vs_CC_N, 15)
ggplot(head(enrichment_table_acetylation_loss_CC_H_vs_CC_N, 15), aes(fold_enrichment, reorder(description, p_adjust),
                        size=observed_region_hits, color=-log10(p_adjust))) + geom_point() + scale_color_viridis_c()
```

Plotting H3K27ac signal in regions that lose H3K27ac in CC_H compared to CC_N 

```{r}
plotEnrichedHeatmaps(extract_signal(acetylation_loss_CC_H_vs_CC_N, tracks), trim=0.95, multiScale = FALSE, axis_name = "0")
```

```{r}
print(paste0("Mean log2FC among significant differentially acetylated regions: ", mean(acetylation_CC_H_vs_CC_N$logFC)))
print(paste0("Mean log2FC among H3K27ac-gain regions: ", mean(acetylation_gain_CC_H_vs_CC_N$logFC)))
print(paste0("Mean log2FC among H3K27ac-loss regions: ", mean(acetylation_loss_CC_H_vs_CC_N$logFC)))
```

H3K27ac-loss regions are stronger acetylation regions overall, so when they lose signal, the absolute visual decrease is large. H3K27ac-gain regions are lower-signal regions overall, so even if their relative logFC is positive, the absolute increase is visually modest.

# Annotating differentially acetylated regions

```{r}
invisible(capture.output({
  anno_acetylation_gain <- annotatePeak(acetylation_gain_CC_H_vs_CC_N, tssRegion=c(-2000,2000), TxDb=TxDb.Mmusculus.UCSC.mm10.knownGene, annoDb="org.Mm.eg.db")
  anno_acetylation_loss <- annotatePeak(acetylation_loss_CC_H_vs_CC_N, tssRegion=c(-2000,2000), TxDb=TxDb.Mmusculus.UCSC.mm10.knownGene, annoDb="org.Mm.eg.db")
}))

plotAnnoBar(c(anno_acetylation_gain, anno_acetylation_loss), title="Genomic Annotation Distribution (acetylation gain vs loss)")
```





















upset plots, overlaps acetylation between conditions to see if there are dynamic changes, acetylation with accessibility
connect all three parts, coherent story
interpret volcano plots cch vs ctrlh !!!!!!!!!!
stats!!!!!!!!!!!!!!!!! for overlaps heatmap
1st plot in H3K27ac signal around enhancer regions
maybe just nkfb spi1 and klf4