---
title: "part1"
output: html_document
date: "2026-06-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread)      # read alignment
  library(rtracklayer)   # read/write genomic file formats
  library(Biostrings)    # handling sequences
  library(Rfastp)        # read QC/trimming
  library(epiwraps)
})
ah <- AnnotationHub()
```

```{r}
#setwd("/Users/marco/Desktop/epigenHypoxia/")
```

### Peak calling using MACS from .bam files
```{bash engine.opts='-l'}

# 1. Extract everything - comment out if you've already extracted it
tar -xvf hypoxia_data.tar

# 2. Setup output directories
mkdir -p peaks/atac
mkdir -p peaks/cutnrun

# 3. Define sample groups based on your GSE300330 metadata
atac_samples=(SRR31003492 SRR31003493 SRR31003494 SRR31003495 SRR31003496 SRR31003497 SRR31003498 SRR31003499)
cnr_samples=(SRR34076314 SRR34076315 SRR34076316 SRR34076317 SRR34076318 SRR34076319 SRR34076320 SRR34076321 SRR34076322 SRR34076323 SRR34076324 SRR34076325)

# 4. Run Peak Calling for ATAC-seq (with Tn5 shifting)
echo "Starting ATAC-seq peak calling..."
for sample in "${atac_samples[@]}"; do
    echo "Processing ATAC: $sample"
    macs3 callpeak -t ATAC/${sample}.bam \
      -f BAMPE \
      -n ${sample}_ATAC \
      --outdir peaks/atac \
      -g hs \
      --keep-dup all \
      --nomodel --shift -100 --extsize 200 \
      -B
done

# 5. Run Peak Calling for CUT&RUN (Standard narrow peaks)
echo "Starting CUT&RUN peak calling..."
for sample in "${cnr_samples[@]}"; do
    echo "Processing CUT&RUN: $sample"
    macs3 callpeak -t CnR/${sample}.bam \
      -f BAMPE \
      -n ${sample}_CnR \
      --outdir peaks/cutnrun \
      -g hs \
      --keep-dup all \
      -B
done

echo "Peak calling complete. All outputs are in the 'peaks/' directory."
```

### Quality Control
```{bash engine.opts='-l'}
# 1. Run samtools flagstat on all ATAC files
mkdir -p qc_reports
for bam in ATAC/*.bam; do
    samtools flagstat "$bam" > "qc_reports/$(basename "$bam").flagstat"
done

# 2. Run samtools flagstat on all CnR files
for bam in CnR/*.bam; do
    samtools flagstat "$bam" > "qc_reports/$(basename "$bam").flagstat"
done

# 3. Compile everything into one visual report
multiqc qc_reports/ -o qc_final_report/
```

### How to filter bad quality aligned reads:

The standard "fine-tuning" steps for a BAM file include:
1) Filter by Mapping Quality (MAPQ): You throw away reads that map poorly or to multiple places in the genome (usually keeping only reads with a score of MAPQ >= 30).
2) Remove Mitochondrial Reads (Crucial for ATAC-seq): You physically strip out all reads mapping to chromosome M (chrM), as they represent junk background noise.
3) Remove PCR Duplicates: You mark and remove identical technical copies generated during library prep.

