---
title: "part2"
output: html_document
date: "2026-06-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/marco/Desktop/epigenHypoxia/")
```

## Differential Accessibility Analysis with edgeR
```{r}
library(edgeR)

count_matrix <- readRDS("peaks/atac_count_matrix.rds")
cat("Total consensus peaks:", nrow(count_matrix), "\n")
consensus_peaks
count_matrix
# Define groups based on the column names of atac_count_matrix that are: Ctlr_N_1, Ctlr_N_2, Ctlr_H_1, Ctlr_H_2, CC_N_1, CC_N_2, CC_H_1, CC_H_2
groups <- factor(c("Ctlr_N", "Ctlr_N", "Ctlr_H", "Ctlr_H", "CC_N", "CC_N", "CC_H", "CC_H"))

# Initialize edgeR object, filter background noise, and normalize
y <- DGEList(counts = count_matrix, group = groups)
keep <- filterByExpr(y)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
cat("Total tested peaks:", length(keep), "\n")

# Set up the statistical design
design <- model.matrix(~0 + groups)
colnames(design) <- levels(groups)

# Estimate variance (dispersion) and fit the model
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design) # Quasi-Likelihood Negative Binomial

# Define your exact 2x2 experimental questions
my_contrasts <- makeContrasts(   
  Ctlr_H_vs_Ctrl_N = Ctlr_H - Ctlr_N, # Baseline hypoxia response -> log2(Ctlr_H) - log2(Ctlr_N) = log2(Ctlr_H/Ctlr_N) = log2FC
  CC_N_vs_Ctlr_N  = CC_N - Ctlr_N,   # Baseline co-culture response
  CC_H_vs_CC_N   = CC_H - CC_N,     # Hypoxia effect during co-culture
  CC_H_vs_Ctrl_H   = CC_H - Ctlr_H,   # Co-culture effect during hypoxia
  levels = design
)
#Positive logFC: The peak opened up in Hypoxia.
#Negative logFC: The peak closed down in Hypoxia.
#logFC near 0: No change between the two conditions.

# Run the tests for all 4 contrasts
test1 <- glmQLFTest(fit, contrast=my_contrasts[,"Ctlr_H_vs_Ctrl_N"])
test2 <- glmQLFTest(fit, contrast=my_contrasts[,"CC_N_vs_Ctlr_N"])
test3 <- glmQLFTest(fit, contrast=my_contrasts[,"CC_H_vs_CC_N"])
test4 <- glmQLFTest(fit, contrast=my_contrasts[,"CC_H_vs_Ctrl_H"])
# Extract the significant tables for each
diff_Ctlr_H_vs_Ctrl_N <- topTags(test1, n=Inf, p.value=0.05)$table
diff_CC_N_vs_Ctlr_N <- topTags(test2, n=Inf, p.value=0.05)$table
diff_CC_H_vs_CC_N <- topTags(test3, n=Inf, p.value=0.05)$table
diff_CC_H_vs_Ctrl_H <- topTags(test4, n=Inf, p.value=0.05)$table

cat("Significant DA peaks in Ctlr_H_vs_Ctrl_N:", nrow(diff_Ctlr_H_vs_Ctrl_N), "\n")
cat("Significant DA peaks in CC_N_vs_Ctlr_N:", nrow(diff_CC_N_vs_Ctlr_N), "\n")
cat("Significant DA peaks in CC_H_vs_CC_N:", nrow(diff_CC_H_vs_CC_N), "\n")
cat("Significant DA peaks in CC_H_vs_Ctrl_H:", nrow(diff_CC_H_vs_Ctrl_H), "\n")

#head(diff_Ctlr_H_vs_Ctrl_N)
```

## Plot volcano plot of DA peaks
```{r}
library(ggplot2)

plot_volcano <- function(df, plot_title) {
  # Categorize each peak
  df$Significance <- "Not Significant"
  df$Significance[df$FDR < 0.05 & df$logFC > 1] <- "Up"
  df$Significance[df$FDR < 0.05 & df$logFC < -1] <- "Down"
  
  # Plotting volcano
  p <- ggplot(df, aes(x = logFC, y = -log10(FDR), color = Significance)) +
    geom_point(alpha = 0.6, size = 1.2) +
    scale_color_manual(values = c("Down" = "blue", "Not Significant" = "gray", "Up" = "red")) +
    theme_minimal() +
    labs(title = plot_title, x = "Log2 Fold Change", y = "-Log10 FDR") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed")
  
  print(p)
  return(df)
}

diff_Ctlr_H_vs_Ctrl_N <- plot_volcano(diff_Ctlr_H_vs_Ctrl_N, "Hypoxia vs Normoxia in monoculture")
diff_CC_H_vs_CC_N <- plot_volcano(diff_CC_H_vs_CC_N, "Hypoxia vs Normoxia in co-culture")
diff_CC_N_vs_Ctlr_N <- plot_volcano(diff_CC_N_vs_Ctlr_N, "Co vs monocolture in Normoxia")
diff_CC_H_vs_Ctrl_H <- plot_volcano(diff_CC_H_vs_Ctrl_H, "Co vs monocolture in Hypoxia")

```

## Plot Veen diagram (absolute values)
```{r}
library(limma)

# passing the original DGELRT objects
results_mat <- cbind(
  Ctlr_H_vs_Ctrl_N = decideTests(test1, p.value = 0.05, lfc = 1),
  CC_N_vs_Ctlr_N     = decideTests(test2, p.value = 0.05, lfc = 1),
  CC_H_vs_CC_N   = decideTests(test3, p.value = 0.05, lfc = 1),
  CC_H_vs_Ctrl_H   = decideTests(test4, p.value = 0.05, lfc = 1)
)

par(mar = c(4, 4, 8, 4), cex = 0.7)  # Increase top margin, shrink text
vennDiagram(results_mat, 
            include = c("up", "down"), 
            counts.col = c("blue", "red"),
            circle.col = c("green", "purple", "black", "red"),
            main = "All conditions overlap", 
            cex.main = 2.5) 

# -- COMBINATIONS

# Does hypoxia open/close the same peaks in monoculture vs. co-culture?
results_hypoxia <- cbind(
  Hypoxia_Monoculture = decideTests(test1, p.value = 0.05, lfc = 1), # Ctlr_H vs Ctlr_N
  Hypoxia_CoCulture   = decideTests(test3, p.value = 0.05, lfc = 1)  # CC_H vs CC_N
)

# Does being in co-culture change accessibility in the same way under normoxia AND hypoxia?
results_coculture <- cbind(
  CoCulture_Normoxia = decideTests(test2, p.value = 0.05, lfc = 1), # CC_N vs Ctlr_N
  CoCulture_Hypoxia  = decideTests(test4, p.value = 0.05, lfc = 1)  # CC_H vs Ctlr_H
)

# Are there peaks that respond to BOTH hypoxia AND co-culture simultaneously?
results_synergy <- cbind(
  Hypoxia_Effect      = decideTests(test1, p.value = 0.05, lfc = 1), # Ctlr_H vs Ctlr_N
  CoCulture_Hypoxia   = decideTests(test4, p.value = 0.05, lfc = 1)  # CC_H vs Ctlr_H
)

par(mar = c(4, 4, 8, 4), cex = 0.7) 

# 1. Hypoxia Response
vennDiagram(results_hypoxia, 
            include = c("up", "down"), 
            counts.col = c("blue", "red"),
            names = c("Hypoxia_Monoculture", "Hypoxia_CoCulture"),
            main = "Hypoxia Response", 
            cex.main = 2.5)

# 2. Co-culture Effect
vennDiagram(results_coculture, 
            include = c("up", "down"), 
            counts.col = c("blue", "red"),
            names = c("CoCulture_Normoxia", "CoCulture_Hypoxia"), 
            main = "Co-culture Effect", 
            cex.main = 2.5)

# 3. Synergy
vennDiagram(results_synergy, 
            include = c("up", "down"), 
            counts.col = c("blue", "red"),
            names = c("Hypoxia_Effect", "CoCulture_Hypoxia"), 
            main = "Synergy Analysis", 
            cex.main = 2.5)
```

#### In these plots, **blue numbers (up)** represent chromatin regions that significantly opened (positive logFC), and **red numbers (down)** represent regions that significantly closed (negative logFC). The large numbers at the bottom right indicate peaks that did not significantly change in either condition.

### **1. Hypoxia Response**

**Question Answered:** How does low oxygen change microglial chromatin, and does the presence of glioma cells change that response?

* **Intersection (554 up, 246 down):** The "core hypoxia signature." These peaks consistently open or close in response to hypoxia, whether the microglia are alone or with glioma cells.
* **Left Circle Unique (610 up, 479 down):** Peaks that only change when microglia experience hypoxia in isolation.
* **Right Circle Unique (1833 up, 838 down):** Peaks that only change when microglia experience hypoxia *while interacting with glioma*.
* **Conclusion:** Co-culture massively amplifies the hypoxia response. Glioma cells prime the microglia to undergo a much broader and more aggressive chromatin remodeling under hypoxic stress than they would alone.

### **2. Co-culture Effect**

**Question Answered:** How does physical interaction with glioma cells alter microglial chromatin, and does oxygen level dictate this interaction?

* **Intersection (763 up, 148 down):** The "core glioma interaction signature." These peaks open or close simply because the tumor is present, regardless of oxygen availability.
* **Left Circle Unique (114 up, 226 down):** Changes triggered by glioma contact strictly in normal oxygen (normoxia).
* **Right Circle Unique (2527 up, 1240 down):** Changes triggered by glioma contact strictly under hypoxic stress.
* **Conclusion:** Hypoxia acts as a powerful catalyst for tumor-microglia communication. Interaction with the tumor in a hypoxic environment unlocks thousands of chromatin regions that remain entirely dormant during interaction in a well-oxygenated environment.

### **3. Synergy Analysis**

**Question Answered:** Are the peaks driven purely by hypoxia the exact same peaks driven by adding glioma cells into a hypoxic environment?

* **Intersection (33 up, 23 down):** Almost zero overlap.
* **Conclusion:** The combination of hypoxia and glioma interaction does not just "double down" on the standard hypoxia response. Adding glioma cells into a hypoxic environment activates an entirely distinct epigenomic program that is completely different from the baseline hypoxia response.

## Genomic Annotation Distribution Across Conditions
```{r}
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

# Load consensus peaks coordinates
consensus_peaks <- readRDS("peaks/atac_consensus.rds")

# Extract the coordinate features for each contrast (FDR < 0.05, |logFC| > 1)
get_granges <- function(diff_table, consensus) {
  
  # Filter for significant rows
  sig_peaks <- rownames(diff_table)[abs(diff_table$logFC) > 1 & diff_table$FDR < 0.05]
  
  # Map back to GRanges object
  if ("name" %in% colnames(mcols(consensus))) {
    return(consensus[consensus$name %in% sig_peaks])
  } else {
    return(consensus[names(consensus) %in% sig_peaks])
  }
}

# Generate GRanges for comparisons
gr_CC   <- get_granges(diff_CC_N_vs_Ctlr_N, consensus_peaks)
gr_Hyp  <- get_granges(diff_Ctlr_H_vs_Ctrl_N, consensus_peaks)
gr_CCHyp <- get_granges(diff_CC_H_vs_CC_N, consensus_peaks)

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

# Annotate genomic features
anno_CC    <- annotatePeak(gr_CC, tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Mm.eg.db")
anno_Hyp   <- annotatePeak(gr_Hyp, tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Mm.eg.db")
anno_CCHyp <- annotatePeak(gr_CCHyp, tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Mm.eg.db")

# Plot combined distributions to recreate Figure 5B
plotAnnoBar(list(CC=anno_CC, CC_Hyp=anno_CCHyp, Hyp=anno_Hyp), 
            title="Genomic Annotation Distribution Across Conditions")
```


### Finding SYNERGISTIC-peaks associated genes
```{r}
# Get the row index names for the overlapping peaks
up_synergy_ids <- rownames(results_synergy)[results_synergy[, 1] == 1 & results_synergy[, 2] == 1]
down_synergy_ids <- rownames(results_synergy)[results_synergy[, 1] == -1 & results_synergy[, 2] == -1]

# Map IDs back to your consensus genomic coordinates object
gr_syn_up   <- consensus_peaks[names(consensus_peaks) %in% up_synergy_ids]
gr_syn_down <- consensus_peaks[names(consensus_peaks) %in% down_synergy_ids]

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

anno_syn_up   <- annotatePeak(gr_syn_up, tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Mm.eg.db")
anno_syn_down <- annotatePeak(gr_syn_down, tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Mm.eg.db")

# Convert annotations to standard data frames
df_syn_up   <- as.data.frame(anno_syn_up)
df_syn_down <- as.data.frame(anno_syn_down)

# Extract unique gene symbols 
genes_up   <- unique(df_syn_up$SYMBOL)
genes_down <- unique(df_syn_down$SYMBOL)

# Print clean lists to your console
cat("--- GENES NEXT TO SYNERGISTIC OPENING PEAKS (33) ---\n")
print(genes_up)

cat("\n--- GENES NEXT TO SYNERGISTIC CLOSING PEAKS (23) ---\n")
print(genes_down)
```

## Venn diagram - considering only promoters peak
```{r}
library(ChIPseeker)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(org.Mm.eg.db)

# Annotate ALL consensus peaks first
txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene
anno_all <- annotatePeak(consensus_peaks, tssRegion=c(-2000, 2000), TxDb=txdb, annoDb="org.Mm.eg.db")
df_anno <- as.data.frame(anno_all)

# Extract IDs for peaks specifically located at Promoters
peak_ids <- if("name" %in% colnames(df_anno)) df_anno$name else rownames(df_anno)
promoter_ids <- peak_ids[grepl("Promoter", df_anno$annotation)]

# 1. Add the missing prefix to match edgeR's naming format
correct_promoter_ids <- paste0("ATAC_IDR_peak_", promoter_ids)

# Quick safety check print
cat("Sample of corrected IDs:", head(correct_promoter_ids), "\n")

# 2. Subset your tests using the corrected names
library(limma)
res_CC    <- decideTests(test2[rownames(test2) %in% correct_promoter_ids, ], p.value = 0.05, lfc = 1)
res_Hyp   <- decideTests(test1[rownames(test1) %in% correct_promoter_ids, ], p.value = 0.05, lfc = 1)
res_CCHyp <- decideTests(test3[rownames(test3) %in% correct_promoter_ids, ], p.value = 0.05, lfc = 1)

# 3. Combine into the final 3-way matrix
results_3way <- cbind(
  CC     = res_CC,
  Hyp    = res_Hyp,
  CC_Hyp = res_CCHyp
)

# 4. Re-plot the Venn Diagram (The numbers should appear now)
par(mar = c(4, 4, 8, 4), cex = 0.7)
vennDiagram(results_3way, 
            include = c("up", "down"), 
            counts.col = c("blue", "red"),
            main = "Promoter Peaks: 3-Way Interaction", 
            cex.main = 2.0)
```
### Finding peaks thanks to the unique combination of CC and Hyp
```{r}
# 1. Isolate the exclusive peak row names using column positions (1=CC, 2=Hyp, 3=CC_Hyp)
exc_up_idx <- which(results_3way[, 1] == 0 & 
                    results_3way[, 2] == 0 & 
                    results_3way[, 3] == 1)

exc_down_idx <- which(results_3way[, 1] == 0 & 
                      results_3way[, 2] == 0 & 
                      results_3way[, 3] == -1)

exc_up_peaks <- rownames(results_3way)[exc_up_idx]
exc_down_peaks <- rownames(results_3way)[exc_down_idx]

# 2. Build a matching id column in your original annotation dataframe
df_anno$fixed_id <- paste0("ATAC_IDR_peak_", rownames(df_anno))

# 3. Pull the exclusive gene sets
genes_up_exclusive   <- unique(df_anno$SYMBOL[df_anno$fixed_id %in% exc_up_peaks])
genes_down_exclusive <- unique(df_anno$SYMBOL[df_anno$fixed_id %in% exc_down_peaks])

# 4. Print clean lists to your console
cat("\n--- GENES WITH EXCLUSIVELY OPENED PROMOTERS ---\n")
print(head(sort(genes_up_exclusive), 50))

cat("\n--- GENES WITH EXCLUSIVELY CLOSED PROMOTERS ---\n")
print(head(sort(genes_down_exclusive), 50))
```

Promoter peaks** (within ±2000 bp of a Transcription Start Site) and analyzed them using a **3-way Venn diagram**.

Instead of comparing just two groups, they looked at the intersection of three distinct operational contrasts:

1. 
**`CC`**: Co-culture effect under normoxia (`CC_N` vs `Ctrl_N`).

2. 
**`Hyp`**: Hypoxia effect in monoculture (`Ctrl_H` vs `Ctrl_N`).

3. 
**`CC+Hyp`**: Hypoxia effect during co-culture (`CC_H` vs `CC_N`).

---

### How the Paper Calculated the Opening Peaks (Figure 5C)

By mapping the 3-way intersection of opening promoter regions, they found a total of **1,680 peaks** that opened up in the hypoxic co-culture (`CC+Hyp`) condition:

* 
**1,515 peaks (The Specific Core):** These were completely unique to the `CC+Hyp` circle. They do not open when microglia are exposed to tumor cells alone under normal oxygen, nor do they open when exposed to hypoxia alone in a monoculture. They strictly require both hits simultaneously.


* 
**109 peaks (The Hypoxia Overlap):** These peaks opened up in `CC+Hyp` but also overlapped with the baseline hypoxia-alone (`Hyp`) circle ($83 + 26$ shared peaks in their 3-way plot).


---

### How the Paper Calculated the Closing Peaks (Figure 5D)

They applied the exact same 3-way partitioning to the **1,439 promoter peaks** that closed down during hypoxic co-culture:

* 
**951 peaks:** Were completely unique to the combined `CC+Hyp` circle.


* 
**483 peaks:** Overlapped with baseline hypoxia-alone closing events ($480 + 3$ shared peaks in their plot).

---

### The Biological Conclusion of This Analysis

This specific mathematical breakdown allowed the authors to discover a major dual mechanism in brain tumors:

* 
**Hypoxia alone is primarily a silencer:** When microglia are by themselves, low oxygen mostly acts as a genomic brake, actively closing down promoter regions.


* 
**The tumor subverts the architecture:** The moment you introduce glioma cells into that exact same low-oxygen space, they completely rewrite the rules. Instead of letting those switches turn off, the tumor-microglia interaction overrides the system, forcing **1,515 entirely new gene promoters** to violently open up and drive the aggressive, immunosuppressive state.


---- difference mine and paper code:
Your code and the paper are answering **different** questions.

Here is exactly why your comparison is different, why it is still useful, and the roadmap you should follow for the clearest biological picture.

### The Difference in "Synergy"

**Your Synergy (The "Additive" Effect):**
Your code asks: *Which peaks are pushed by hypoxia alone, and then pushed **again** by the tumor?* * `Ctlr_H vs Ctlr_N`: Does hypoxia change this?

* `CC_H vs Ctlr_H`: Does adding the tumor change it further?
* **Why you need it:** This identifies genes that are highly sensitive to *both* distinct stressors. It finds the "double-hit" genes.

**The Paper's Synergy (The "Exclusive" Effect):**
The paper asks: *Which peaks **only** change when both hypoxia and the tumor are present at the same time?*

* They compared three different states simultaneously to find peaks that were completely dormant in monoculture (even under hypoxia) but violently opened in co-culture under hypoxia.

---

### What You Did vs. What the Paper Did

| Feature | Your Analysis | The Paper's Analysis |
| --- | --- | --- |
| **Genomic Scope** | **Whole Genome:** Analyzed all peaks (promoters, introns, intergenic). | **Promoters Only:** Filtered strictly for peaks $\pm 2000$ bp from a TSS first. |
| **Math Model** | **2-Way Intersection:** Compared two contrasts at a time. | **3-Way Intersection:** Compared three contrasts simultaneously. |
| **Core Question** | Which peaks are compounded by both stressors? | Which promoters are exclusively unlocked by the dual-environment? |

# Over-Representation Analysis (ORA) with clusterProfiler and enrichGO - BACKGROUND important -> all consensus peaks
```{r}
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(ggplot2)

length(genes_up_exclusive)
length(genes_down_exclusive)

# Annotate the ENTIRE consensus peak set to capture the full ATAC "universe"
anno_consensus <- annotatePeak(consensus_peaks, 
                               tssRegion = c(-2000, 2000), 
                               TxDb = txdb, 
                               annoDb = "org.Mm.eg.db")
# Extract the data frame
df_anno_consensus <- as.data.frame(anno_consensus)
# Create your background universe (all unique gene symbols detected in the ATAC-seq experiment)
all_atac_genes <- unique(df_anno_consensus$SYMBOL)
all_atac_genes <- all_atac_genes[!is.na(all_atac_genes)] # Remove NAs

# 2. Run GO Enrichment (ORA) for UP genes
go_up <- enrichGO(
  gene          = genes_up_exclusive,
  universe      = all_atac_genes,      # background
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",  # Your vectors contain gene symbols
  ont           = "ALL",      # "BP" = Biological Process (can also use "MF", "CC", or "ALL")
  pAdjustMethod = "BH",      # Benjamini-Hochberg for FDR
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# 4. Run the GO enrichment using this universe
go_down <- enrichGO(
  gene          = genes_down_exclusive,
  universe      = all_atac_genes,      # background
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "ALL",               # Checks BP, MF, and CC or ALL
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# Plot the results
dotplot(go_up, showCategory = 15, font.size = 9) + 
  ggtitle("GO terms: Exclusively Opened Promoters (UP)") +
  theme(plot.title = element_text(size = 11, face = "bold"))

dotplot(go_down, showCategory = 15, font.size = 9) + 
  ggplot2::ggtitle("GO terms: Exclusively Closed Promoters (DOWN)") +
  ggplot2::theme(plot.title = ggplot2::element_text(size = 11, face = "bold"))
```

### how the paper did, just copy pasted from gemini, i didn't check, i'm going to bed, tomorrow i'll clean up the analysis and find out what's the best 
```{r}
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(ggplot2)

# 1. Create a matching ID column in your global annotation dataframe
peak_ids <- if("name" %in% colnames(df_anno)) df_anno$name else rownames(df_anno)
df_anno$peak_id <- paste0("ATAC_IDR_peak_", peak_ids)

# 2. Extract significant edgeR results for the Hypoxic Co-culture effect (CC_H vs CC_N)
sig_CC_Hyp <- diff_CC_H_vs_CC_N[diff_CC_H_vs_CC_N$FDR < 0.05 & abs(diff_CC_H_vs_CC_N$logFC) > 1, ]
sig_CC_Hyp$peak_id <- rownames(sig_CC_Hyp)

# 3. Merge edgeR stats with genomic annotations
merged_CC_Hyp <- merge(sig_CC_Hyp, df_anno, by="peak_id")

# 4. Filter strictly for GENIC peaks (Promoter, Intron, Exon)
# This removes "Distal Intergenic" and "Downstream" peaks
genic_CC_Hyp <- merged_CC_Hyp[grepl("Promoter|Intron|Exon", merged_CC_Hyp$annotation), ]

# 5. Extract unique gene symbols for UP and DOWN peaks
genes_genic_up <- unique(genic_CC_Hyp$SYMBOL[genic_CC_Hyp$logFC > 1])
genes_genic_down <- unique(genic_CC_Hyp$SYMBOL[genic_CC_Hyp$logFC < -1])

cat("Genic UP genes:", length(genes_genic_up), "\n")
cat("Genic DOWN genes:", length(genes_genic_down), "\n")

# 6. Run enrichGO WITHOUT the universe parameter to match the paper's stats
go_genic_up <- enrichGO(
  gene          = genes_genic_up,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",      # Biological Process
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

go_genic_down <- enrichGO(
  gene          = genes_genic_down,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",      # Biological Process
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

# 7. Plot the results
dotplot(go_genic_up, showCategory = 15, font.size = 9) + 
  ggtitle("GO terms: Upregulated Genic Peaks (Hypoxic CC)") +
  theme(plot.title = element_text(size = 11, face = "bold"))

dotplot(go_genic_down, showCategory = 15, font.size = 9) + 
  ggtitle("GO terms: Downregulated Genic Peaks (Hypoxic CC)") +
  theme(plot.title = element_text(size = 11, face = "bold"))
```

# So the paper did wrong because they didn't use a background, are you sure gemini? I should check in the paper...

To perfectly mirror the paper and recreate Figures 5H and 5I, you need to change two specific things about your current setup:
Scope: The authors did not use the 3-way exclusive promoters for this specific GO analysis. They used Genic peaks (which includes Promoters + Introns + Exons) specifically from the Hypoxic Co-culture contrast.
Background: They did not apply a custom universe. They let the tool default to the whole mouse genome, which inflates significance but is necessary to replicate their exact plots.



## Average .bw files for IGV visualization (for the image I uploaded)
```{r, eval=FALSE}
library(rtracklayer)
library(GenomicRanges)

# Set your directory
setwd("~/Desktop/epigenHypoxia/filtered_bams")

average_bigwigs <- function(file1, file2, output_file) {
  if (!file.exists(file1)) stop(paste("Missing file:", file1))
  if (!file.exists(file2)) stop(paste("Missing file:", file2))
  
  message("Averaging: ", file1, " and ", file2)
  
  # 1. Import bigWig files as genomic ranges
  bw1 <- import(file1)
  bw2 <- import(file2)
  
  # 2. Convert to base-by-base tracks using the 'score' column
  cov1 <- coverage(bw1, weight="score")
  cov2 <- coverage(bw2, weight="score")
  
  # 3. Ensure both tracks share the exact same chromosomes
  common_chroms <- intersect(names(cov1), names(cov2))
  cov1 <- cov1[common_chroms]
  cov2 <- cov2[common_chroms]
  
  # 4. Calculate the mathematical mean across the genome
  avg_cov <- (cov1 + cov2) / 2
  
  # 5. Export back to a clean bigWig file
  export.bw(avg_cov, output_file)
  message("Saved: ", output_file, "\n")
}

# Run the averaging operations
# 1. ATAC-seq Controls
average_bigwigs("ATAC_Ctlr_N_1.bw", "ATAC_Ctlr_N_2.bw", "ATAC_Ctlr_N_averaged.bw")
average_bigwigs("ATAC_Ctlr_H_1.bw", "ATAC_Ctlr_H_2.bw", "ATAC_Ctlr_H_averaged.bw")

# 2. ATAC-seq Co-Cultures
average_bigwigs("ATAC_CC_N_1.bw", "ATAC_CC_N_2.bw", "ATAC_CC_N_averaged.bw")
average_bigwigs("ATAC_CC_H_1.bw", "ATAC_CC_H_2.bw", "ATAC_CC_H_averaged.bw")

# 3. H3K27ac CUT&RUN
average_bigwigs("SRR34076314_H3K27ac.bw", "SRR34076315_H3K27ac.bw", "H3K27ac_averaged.bw")

```

