---
title: "part1"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2026-06-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set( warning = FALSE, message = FALSE)
```

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(Rsubread)      # read alignment
  library(rtracklayer)   # read/write genomic file formats
  library(Biostrings)    # handling sequences
  library(Rfastp)        # read QC/trimming
  library(epiwraps)
})
```

## Pre-processing

### Load aligned .bam files of ATACseq and H3K27ac CUT&RUN
```{bash engine.opts='-l', eval=FALSE}
tar -xvf hypoxia_data.tar
```

### Quality Control aligned reads (.bam files)
```{bash engine.opts='-l', eval= FALSE}
mkdir -p qc_reports
mkdir -p qc_reports/idxstats

for bam in ATAC/*.bam CnR/*.bam; do
    samtools flagstat "$bam" > "qc_reports/$(basename "$bam").flagstat"
    samtools idxstats "$bam" > "qc_reports/idxstats/$(basename "$bam").idxstats"
done

multiqc qc_reports/ -o qc_final_report/
```

### Filtering low quality .bam reads
```{bash engine.opts='-l', eval= FALSE}
mkdir -p filtered_bams qc_reports/dup_metrics

for bam in ATAC/*.bam CnR/*.bam; do
    base=$(basename "$bam" .bam)
    samtools view -h -q 30 "$bam" | awk '$1 ~ /^@/ || ($3 != "chrM" && $3 != "MT")' | samtools view -b > "filtered_bams/${base}_temp1.bam"
    samtools sort -n "filtered_bams/${base}_temp1.bam" | samtools fixmate -m - - | samtools sort - | samtools markdup -r - "filtered_bams/${base}_final.bam"
    samtools index "filtered_bams/${base}_final.bam"
    rm "filtered_bams/${base}_temp1.bam"
done
```

### Quality Control on filtered aligned reads (_final.bam files)
```{bash engine.opts='-l', eval= FALSE}
mkdir -p qc_reports_filt/idxstats

for bam in filtered_bams/*.bam; do
    samtools flagstat "$bam" > "qc_reports_filt/$(basename "$bam").flagstat"
    samtools idxstats "$bam" > "qc_reports_filt/idxstats/$(basename "$bam").idxstats"
done

multiqc qc_reports_filt/ -o qc_final_report_filt/
```

### Peak calling and bamCoverage
```{bash engine.opts='-l', eval=FALSE}
mkdir -p peaks/atac
mkdir -p peaks/cutnrun

atac_samples=(SRR31003492 SRR31003493 SRR31003494 SRR31003495 SRR31003496 SRR31003497 SRR31003498 SRR31003499)
cnr_samples=(SRR34076314 SRR34076315 SRR34076316 SRR34076317 SRR34076318 SRR34076319 SRR34076320 SRR34076321 SRR34076322 SRR34076324)

# Run Peak Calling for ATAC-seq
for sample in "${atac_samples[@]}"; do
    echo "Processing ATAC: $sample"
    macs3 callpeak -t "filtered_bams/${sample}_final.bam" \
      -f BAMPE \
      -n "${sample}_ATAC" \
      --outdir peaks/atac \
      -g mm \
      --nomodel \
      --shift 0 \
      -q 0.01
done

# Run Peak Calling for CUT&RUN H3K27ac
for sample in "${cnr_samples[@]}"; do
    echo "Processing CUT&RUN: $sample"
    macs3 callpeak -t "filtered_bams/${sample}_final.bam" \
      -f BAMPE \
      -n "${sample}_H3K27ac" \
      --outdir peaks/cutnrun \
      -g mm \
      --broad \
      --broad-cutoff 0.1 \
      -q 0.05
done

# Run bamCoverage for ATAC-seq
for sample in "${atac_samples[@]}"; do
    echo "Processing ATAC to bigWig: $sample"
    bamCoverage -b "filtered_bams/${sample}_final.bam" \
      -o "filtered_bams/${sample}_ATAC.bw" \
      --normalizeUsing RPKM \
      -p max
done

# Run bamCoverage for CUT&RUN
for sample in "${cnr_samples[@]}"; do
    echo "Processing CUT&RUN: $sample"
    bamCoverage -b "filtered_bams/${sample}_final.bam" \
      -o "filtered_bams/${sample}_H3K27ac.bw" \
      --normalizeUsing RPKM \
      -p max
done
echo "Peak calling and bigWig conversion complete"
```

## Technical Quality Control
### Quality Control on peak counts of ATACseq (from Peak Calling)
```{r}
library(GenomicRanges)
library(ggplot2)

atac_peak_files <- list.files("peaks/atac", pattern = "narrowPeak$", full.names = TRUE)
atac_peaks_list <- lapply(atac_peak_files, rtracklayer::import)

meta_atac <- read.csv("ATAC/SraRunTable_hypoxia_ATAC.csv")

df_atac <- data.frame(
  Run = gsub("_ATAC.*", "", basename(atac_peak_files)),
  Peak_Counts = lengths(atac_peaks_list),
  stringsAsFactors = FALSE
)
df_atac <- merge(df_atac, meta_atac, by = "Run")
df_atac$Group <- c(
  "Monoculture_normoxia 21% O2"  = "Ctrl_N",
  "Co-culture_normoxia 21% O2"   = "CC_N",
  "Monoculture_hypoxia <0.1% O2" = "Ctrl_H",
  "Co-culture_hypoxia <0.1% O2"  = "CC_H"
)[paste(df_atac$condition, df_atac$treatment, sep = "_")]

stopifnot(!any(is.na(df_atac$Group)))

ggplot(df_atac, aes(x = Run, y = Peak_Counts, color = Group)) +
  geom_point(size = 5) +
      scale_y_continuous(
    breaks = unique(sort(c(seq(0, 50000, 10000), 20000))),
    limits = c(0, 50000) ) +
  scale_color_manual(values = c(
    "Ctrl_N" = "purple", "CC_N"   = "darkgreen",
    "Ctrl_H" = "lightblue", "CC_H"   = "red" )) +
  theme_minimal() +
  labs( title = "ATAC Peak Counts by Group (Condition × Treatment)",
    x = "Sample ID", y = "Number of Peaks", color = "Group" ) + 
  theme( axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"), legend.position = "right")
```

### Quality Control on peak counts of H3K27ac (from Peak Calling)
```{r}
cnr_peak_files <- list.files("peaks/cutnrun", pattern = "broadPeak$", full.names = TRUE)
cnr_peaks_list <- lapply(cnr_peak_files, rtracklayer::import)

meta_cnr <- read.csv("CnR/SraRunTable_hypoxia_CutNRun.csv")

df_cnr <- data.frame(
  Run = gsub("_H3K27ac.*", "", basename(cnr_peak_files)),
  Peak_Counts = lengths(cnr_peaks_list),
  stringsAsFactors = FALSE
)
df_cnr <- merge(df_cnr, meta_cnr, by = "Run")
df_cnr$Group <- c(
  "Monoculture_normoxia 21% O2"  = "Ctrl_N",
  "Co-culture_normoxia 21% O2"   = "CC_N",
  "Monoculture_hypoxia <0.1% O2" = "Ctrl_H",
  "Co-culture_hypoxia <0.1% O2"  = "CC_H"
)[paste(df_cnr$condition, df_cnr$treatment, sep = "_")]

stopifnot(!any(is.na(df_cnr$Group)))

ggplot(df_cnr, aes(x = Run, y = Peak_Counts, color = Group)) +
  geom_point(size = 5) +
  scale_color_manual(values = c(
    "Ctrl_N" = "purple", "CC_N"   = "darkgreen",
    "Ctrl_H" = "lightblue", "CC_H"   = "red" )) +
  theme_minimal() +
  labs( title = "H3K27ac Broad Peak Counts by Group (Condition × Treatment)",
    x = "Sample ID",
    y = "Number of Peaks",
    color = "Group") +
  theme( axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"), legend.position = "right")
    
```

## IDR (Irreproducible Discovery Rate) per condition
```{bash engine.opts='-l', eval= FALSE}
mkdir -p peaks/idr 

idr --samples SRR31003498_ATAC_peaks.narrowPeak SRR31003499_ATAC_peaks.narrowPeak \
    --peak-list cond_Ctlr_N_pooled_peaks.narrowPeak \
    --input-file-type narrowPeak \
    --output-file peaks/idr/cond_Ctlr_N_idr.txt \
    --rank p.value \
    --soft-idr-threshold 0.05 \
    --plot
  
idr --samples SRR31003496_ATAC_peaks.narrowPeak SRR31003497_ATAC_peaks.narrowPeak \
    --peak-list cond_Ctlr_H_pooled_peaks.narrowPeak \
    --input-file-type narrowPeak \
    --output-file peaks/idr/cond_Ctlr_H_idr.txt \
    --rank p.value \
    --soft-idr-threshold 0.05 \
    --plot
    
idr --samples SRR31003494_ATAC_peaks.narrowPeak SRR31003495_ATAC_peaks.narrowPeak \
    --peak-list cond_CC_N_pooled_peaks.narrowPeak \
    --input-file-type narrowPeak \
    --output-file peaks/idr/cond_CC_N_idr.txt \
    --rank p.value \
    --soft-idr-threshold 0.05 \
    --plot
    
idr --samples SRR31003492_ATAC_peaks.narrowPeak SRR31003493_ATAC_peaks.narrowPeak \
    --peak-list cond_CC_H_pooled_peaks.narrowPeak \
    --input-file-type narrowPeak \
    --output-file peaks/idr/cond_CC_H_idr.txt \
    --rank p.value \
    --soft-idr-threshold 0.05 \
    --plot
```

## Filter and Union by IDR to get consensus
```{r}
load_idr_peaks <- function(idr_file_path) {

  idr_data <- read.table(idr_file_path, sep="\t", header=FALSE, stringsAsFactors=FALSE)
  colnames(idr_data)[1:3] <- c("chr", "start", "end")
  colnames(idr_data)[12] <- "globalIDR"
  passed_df <- idr_data[idr_data$globalIDR > 1.301, ]
  
  passed_peaks <- GenomicRanges::makeGRangesFromDataFrame(passed_df, keep.extra.columns=TRUE)
  passed_peaks <- GenomeInfoDb::keepStandardChromosomes(passed_peaks, pruning.mode="coarse")
  
  return(passed_peaks) 
}
cond_Ctlr_N_idr <- load_idr_peaks("peaks/idr/cond_Ctlr_N_idr.txt")
cond_Ctlr_H_idr <- load_idr_peaks("peaks/idr/cond_Ctlr_H_idr.txt")
cond_CC_N_idr <- load_idr_peaks("peaks/idr/cond_CC_N_idr.txt")
cond_CC_H_idr <- load_idr_peaks("peaks/idr/cond_CC_H_idr.txt")

atac_consensus <- GenomicRanges::reduce(c(cond_Ctlr_N_idr, cond_Ctlr_H_idr, cond_CC_N_idr, cond_CC_H_idr))
names(atac_consensus) <- paste0("ATAC_IDR_peak_", seq_along(atac_consensus))
saveRDS(atac_consensus, "peaks/atac_consensus.rds")
```

### PCA on ATACseq (from Peak Calling)
```{r}
peak_matrix <- sapply(atac_peaks_list, function(x) {
  as.integer(overlapsAny(atac_consensus, x))
})
colnames(peak_matrix) <- gsub("_ATAC.*", "", basename(atac_peak_files))

pca_res <- prcomp(t(peak_matrix))
var_pct <- round(100 * (pca_res$sdev^2 / sum(pca_res$sdev^2)), 1)

pca_df <- data.frame(
  Run = colnames(peak_matrix),
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2]
)
pca_df <- merge(pca_df, df_atac, by = "Run")

stopifnot(!any(is.na(pca_df$Group)))

ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 6) +
  scale_color_manual(values = c(
    "Ctrl_N" = "purple",
    "CC_N"   = "darkgreen",
    "Ctrl_H" = "lightblue",
    "CC_H"   = "red" )) + 
  labs(title = "PCA ATACseq",
    x = paste0("PC1: ", var_pct[1], "%"),
    y = paste0("PC2: ", var_pct[2], "%"),
    color = "Group") +
  theme_bw() +
  theme( panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right")
```

### Consensus H3K27ac by Intersection
```{r}
library(GenomeInfoDb)

get_replicated_per_cond <- function(file_paths, min_reps = 2) {
  
  peaks <- lapply(file_paths, rtracklayer::import)
  peaks <- lapply(peaks, function(p) keepStandardChromosomes(p, pruning.mode="coarse"))  
  unioned <- GenomicRanges::reduce(do.call(c, peaks))
  
  overlap_counts <- rowSums(sapply(peaks, function(p) {
    GenomicRanges::countOverlaps(unioned, p) > 0
  }))
  
  return(unioned[overlap_counts >= min_reps])
}
cnr_Ctlr_N <- get_replicated_per_cond(c("peaks/cutnrun/SRR34076321_H3K27ac_peaks.broadPeak", 
                                        "peaks/cutnrun/SRR34076324_H3K27ac_peaks.broadPeak"))
cnr_Ctlr_H <- get_replicated_per_cond(c("peaks/cutnrun/SRR34076319_H3K27ac_peaks.broadPeak", 
                                        "peaks/cutnrun/SRR34076322_H3K27ac_peaks.broadPeak"))
cnr_CC_N <- get_replicated_per_cond(c("peaks/cutnrun/SRR34076317_H3K27ac_peaks.broadPeak", 
                                      "peaks/cutnrun/SRR34076318_H3K27ac_peaks.broadPeak"))
cnr_CC_H <- get_replicated_per_cond(c("peaks/cutnrun/SRR34076314_H3K27ac_peaks.broadPeak", 
                                      "peaks/cutnrun/SRR34076316_H3K27ac_peaks.broadPeak"))

H3K27ac_consensus <- GenomicRanges::reduce(c(cnr_Ctlr_N, cnr_Ctlr_H, cnr_CC_N, cnr_CC_H))
names(H3K27ac_consensus) <- paste0("H3K27ac_peak_", seq_along(H3K27ac_consensus))
saveRDS(H3K27ac_consensus, "peaks/H3K27ac_consensus.rds")
```

### PCA on H3K27ac (from Peak Calling) 
```{r, eval=FALSE}
H3K27ac_peak_files <- list.files("peaks/cutnrun", pattern = "broadPeak$", full.names = TRUE)
H3K27ac_peaks_list <- lapply(H3K27ac_peak_files, rtracklayer::import)

peak_matrix <- sapply(cnr_peaks_list, function(x) {
  as.integer(overlapsAny(H3K27ac_consensus, x))
})
colnames(peak_matrix) <- gsub("_H3K27ac.*", "", basename(H3K27ac_peak_files))

pca_res <- prcomp(t(peak_matrix))
var_pct <- round(100 * (pca_res$sdev^2 / sum(pca_res$sdev^2)), 1)

pca_df <- data.frame(
  Run = colnames(peak_matrix),
  PC1 = pca_res$x[, 1],
  PC2 = pca_res$x[, 2]
)
pca_df <- merge(pca_df, df_cnr, by = "Run")

stopifnot(!any(is.na(pca_df$Group)))

ggplot(pca_df, aes(x = PC1, y = PC2, color = Group)) +
  geom_point(size = 6) +
  scale_color_manual(values = c(
    "Ctrl_N" = "purple",
    "CC_N"   = "darkgreen",
    "Ctrl_H" = "lightblue",
    "CC_H"   = "red" )) +
  labs( title = "PCA H3K27ac",
    x = paste0("PC1: ", var_pct[1], "%"),
    y = paste0("PC2: ", var_pct[2], "%"),
    color = "Group") +
  theme_bw() +
  theme( panel.grid = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right")
```

## Creating atac_count_matrix
```{r, results = 'hide'}
library(Rsubread)

saf <- data.frame(
  GeneID = names(atac_consensus),
  Chr    = as.character(seqnames(atac_consensus)),
  Start  = start(atac_consensus),
  End    = end(atac_consensus),
  Strand = "*"
)

bam_files <- c(
  "filtered_bams/SRR31003498_final.bam", "filtered_bams/SRR31003499_final.bam", # Ctlr_N
  "filtered_bams/SRR31003496_final.bam", "filtered_bams/SRR31003497_final.bam", # Ctlr_H
  "filtered_bams/SRR31003494_final.bam", "filtered_bams/SRR31003495_final.bam", # CC_N
  "filtered_bams/SRR31003492_final.bam", "filtered_bams/SRR31003493_final.bam"  # CC_H
)
counts <- Rsubread::featureCounts(
  files = bam_files,
  annot.ext = saf,
  isPairedEnd = TRUE, 
  requireBothEndsMapped = TRUE,
  nthreads = 4
)
count_matrix <- counts$counts

colnames(count_matrix) <- c(
  "Ctlr_N_1", "Ctlr_N_2",
  "Ctlr_H_1",  "Ctlr_H_2",
  "CC_N_1", "CC_N_2",
  "CC_H_1",  "CC_H_2"
)
saveRDS(count_matrix, "peaks/atac_count_matrix.rds")
```

## Creating H3K27ac_count_matrix
```{r, results = 'hide'}
saf_cnr <- data.frame(
  GeneID = names(H3K27ac_consensus),
  Chr    = as.character(seqnames(H3K27ac_consensus)),
  Start  = start(H3K27ac_consensus),
  End    = end(H3K27ac_consensus),
  Strand = "*"
)
h3k27ac_bam_files <- c(
  "filtered_bams/SRR34076321_final.bam", "filtered_bams/SRR34076324_final.bam", # Ctlr_N
  "filtered_bams/SRR34076319_final.bam", "filtered_bams/SRR34076322_final.bam", # Ctlr_H
  "filtered_bams/SRR34076317_final.bam", "filtered_bams/SRR34076318_final.bam", # CC_N
  "filtered_bams/SRR34076314_final.bam", "filtered_bams/SRR34076316_final.bam"  # CC_H
)
h3k27ac_counts <- Rsubread::featureCounts(
  files = h3k27ac_bam_files,
  annot.ext = saf_cnr,
  isPairedEnd = TRUE, 
  requireBothEndsMapped = TRUE,
  nthreads = 4
)
h3k27ac_count_matrix <- h3k27ac_counts$counts

colnames(h3k27ac_count_matrix) <- c(
  "Ctlr_N_1", "Ctlr_N_2",
  "Ctlr_H_1", "Ctlr_H_2",
  "CC_N_1",   "CC_N_2",
  "CC_H_1",   "CC_H_2"
)

saveRDS(h3k27ac_count_matrix, "peaks/h3k27ac_count_matrix.rds")
```



